---
title: "Musical Ambiguous Figure Data Analysis"

output: html_notebook
---
# Musical Ambiguous Figures Project: Data Analysis

**Author:** Sarah Gates
**Contact** sarahgates22@gmail.com
**Date:** Summer, 2022

## Introduction

This R notebook contains the statistical analysis for my Musical Ambiguous Figures project. In the first section, I perform the statistical analyses on the first dependent variable, ease of hearing ratings. I first xamine the assumptions for parametric statistical analysis (outliers, normality), then perform 


First, I will load the required packages:

```{r}
library(ggplot2)
library(tidyverse)
library(ggpubr)
library(rstatix)
```

# Statistical Analysis: Dependent Variable 1 (Ease of Hearing Ratings)

In this section, I perform the analyses on the first dependent variable. 

```{r}
# loading the data
df <- readfile <- read.csv("data/amb_DV1_fin.csv")

# view the data
View(df)

# Examine the data contents and data types
str(df)
```

## 1. Cursory Exploratory Data Analysis

```{r}
# grouping by the different IVs
df %>%
  group_by(Sonata_Section, Feature, Mod_Type, Expertise_Group) %>%
  get_summary_stats(DV, type = "mean_sd")
```
Following this, I will visualize, using a box plot, the 

```{r}
bxp <- ggboxplot(
  df, x = "Feature", y = "DV",
  color = "Mod_Type", palette = "jco",
  facet.by = "Sonata_Section", short.panel.labs = FALSE
  )
print(bxp)
```
### Outliers

Then I will use 'identify outliers' on the DV, grouped by all levels of the IVs: this shows that several participants have outlier values, including participants 495,473,268, 475, 834,883, 862 and 392. This shows that within each of these subgroups of IVS (e.g. Exposition-Bass-Prinner-Expert), that their individual value for the DV is considered an outlier. 

```{r}
df %>%
  group_by(Sonata_Section, Feature, Mod_Type, Expertise_Group) %>%
  identify_outliers(DV)
```
### Assumption of Normality: Shapiro-Wilk Test

Secondly, we can use the Shapiro-Wilk test to examine whether the DV is normally distributed across levels of the IV:

```{r}
df %>%
  group_by(Sonata_Section, Feature, Mod_Type) %>%
  shapiro_test(DV)
```
#### QQ Plot

Given that all of the p-values are significant here, it is evident that there are sphericity violations in this dataset. As such, Greenhouse-Geisser corrections to p-values will be applied where applicable. 

We can also visualize these violations using QQplots: 

```{r}
ggqqplot(df, "DV", ggtheme = theme_bw()) +
  facet_grid(Sonata_Section ~ Feature ~ Mod_Type, labeller = "label_both")
```
### Assumption of Homogeneity of Variance: Levene Test

We can also test for homogeneity of variance using the Levene Test. This shows that there is homogeneity of variances (p > 0.05) for all factors except for the Exposition version of the Bassline Romanesca schema (p = 0.02)

```{r}
df %>%
  group_by(Sonata_Section, Feature, Mod_Type) %>%
  levene_test(DV ~ Expertise_Group)
```

We can also check the variance for Excerpt Familiarity, which has equal variance (p > 0.05).

```{r}
df %>%
  group_by(Sonata_Section, Feature, Mod_Type) %>%
  levene_test(DV ~ Excerpt_Familiarity)
```

Lastly, we can visualize the mean and standard deviations for the dependent variable, grouped by the levels of factors for reference following the ANOVA (see below).

```{r}
df %>%
  group_by(Sonata_Section, Feature, Mod_Type, Expertise_Group) %>%
  get_summary_stats(DV, type = "mean_sd")
```

## 2. Four-Way Mixed ANOVA: 2*3*2 Factorial Design (Sonata Section, Feature, Modulation Type)

Here I perform a 2*3*2 factorial mixed ANOVA in order to examine the effects of Sonata Section (exposition, recap), attended Feature (bass, soprano, schema), and Modulation Type (Romanesca, Prinner) on ease of hearing ratings. 

```{r}
# Compute RM ANOVA for within and between subjects factors
res.aov.bet <- anova_test(
  data = df, dv = DV, wid = Random_IDs,
  within = c(Sonata_Section, Feature, Mod_Type), between=Expertise_Group
  )
get_anova_table(res.aov.bet, correction="GG")
```

I will also output the greenhouse geisser corrections:

```{r}
res.aov.bet$`Sphericity Corrections`
```


**Summary of Findings**

This shows that there was a significant main effect of expertise (p = 0.007), feature (p < 0.001), and modulation type (p = 0.043), as well as significant two way interactions between expertise group and feature (p = 0.030), expertise group and modulation type (p = 0.024), sonata section and feature (p = 0.006), and feature and modulation type (p = 0.010). 

I will now examine each of these effects using non-parametrics Tukey HSD tests to determine which pair-wise differences the significant effects stem from. 


### Examining the Main Effects

Here, I examine the main effects using non-parametric tuckey hsd tests on each statistically significant main effect. 

**Expertise**

```{r}
# Examine Main Effects

#Expertise

# tukey 
df %>% tukey_hsd(DV ~ Expertise_Group, conf.level = 0.95)

```


```{r}
df %>%
  group_by(Expertise_Group) %>%
  get_summary_stats(DV, type = "mean_sd")
```

This analysis shows that both expert (M = 5.13, sd = 2.21) and intermediate (M = 5.35, sd = 1.60) groups differ from the novice (M = 3.56, sd = 2.17) group. Both expert and intermediate ease of hearing ratings are significantly higher than novice ease of hearing ratings. 

**Feature**

```{r}
# Feature

df %>% tukey_hsd(DV ~ Feature, conf.level = 0.95)

```


```{r}
df %>%
  group_by(Feature) %>%
  get_summary_stats(DV, type = "mean_sd")
```

The main effect of feature stems from the significant difference between bass (M = 5.44, sd = 1.62) and soprano (M = 4.38, sd = 2.08) features. This suggests that the soprano line was the most difficult to hear and attend to.

**Modulation Type**

```{r}
#Mod Type

df %>% tukey_hsd(DV ~ Mod_Type, conf.level = 0.95)

```


```{r}
df %>%
  group_by(Mod_Type) %>%
  get_summary_stats(DV, type = "mean_sd")
```
The main effect of modulation type shows that the Prinner (M = 5.18, sd = 1.95) was easier to hear than the Romanesca (M = 4.65, sd = 2.02). 

### Examining Two-Way Interactions

In this section, I examine the significant two way interactions between expertise group and feature (p = 0.030), expertise group and modulation type (p = 0.024), sonata section and feature (p = 0.006), and feature and modulation type (p = 0.010). 

**Expertise and Feature**

```{r}
# Two Way Interactions

# Expertise Group by Feature

df %>%
  group_by(Expertise_Group) %>%
  tukey_hsd(DV ~ Feature, conf.level = 0.95)
  
```


```{r}
df %>%
  group_by(Expertise_Group, Feature) %>%
  get_summary_stats(DV, type = "mean_sd") 
```

The post-hoc Tukey HSD examining the interaction between expertise group and feature shows that for only the novice group did the bass (M = 4.88, sd = 2.07) and soprano (M = 2.51, sd = 1.62) differ significantly. 

**Expertise and Modulation**

```{r}
# Expertise Group by Modulation Type

df %>%
  group_by(Expertise_Group) %>%
  tukey_hsd(DV ~ Mod_Type, conf.level = 0.95)
```


```{r}
df %>%
  group_by(Expertise_Group, Mod_Type) %>%
  get_summary_stats(DV, type = "mean_sd") 
```
The analysis examining the interaction of expertise group and modulation type shows that only for the expert group did Prinner (M = 6.34, sd = 1.50) and Romanesca (M = 4.02, sd = 2.08) schema differ. 

**Sonata Section and Feature**

```{r}
# Sonata Section by Feature

df %>%
  group_by(Sonata_Section) %>%
  tukey_hsd(DV ~ Feature, conf.level = 0.95)
  
```


```{r}
df %>%
  group_by(Sonata_Section, Feature) %>%
  get_summary_stats(DV, type = "mean_sd")
```
The Tukey HSD test on the interaction between sonata section and feature revealed that only for the recapitulation did soprano (M = 4.17, sd = 2.13) and bass (M = 5.59, sd = 1.44) significantly differ. 

**Feature and Modulation Type**

```{r}
# Feature by Modulation Type

df %>%
  group_by(Feature) %>%
  tukey_hsd(DV ~ Mod_Type, conf.level = 0.95)
  
```

```{r}
df %>%
  group_by(Feature, Mod_Type) %>%
  get_summary_stats(DV, type = "mean_sd")
```
The post hoc Tukey HSD test revealed that for the schema attending condition only did Prinner (M = 5.70, sd = 1.81) and Romanesca (M = 4.15, sd = 2.17) interpretations differ. 

### Examining Effect of Stimuli Presentation Order

In order to examine in the order in which stimuli was presented had an effect on interpretation formation, a second ANOVA including Sonata Order as a between-subjects factor, in addition to expertise group, was completed. This showed however that there was no effect of order of presentation of sonata section on the DV. 

```{r}
res.aov.order <- anova_test(
  data = df, dv = DV, wid = Random_IDs,
  within = c(Sonata_Section, Feature, Mod_Type), between=c(Expertise_Group, Sonata_Order)
  )
get_anova_table(res.aov.order)
```

# Statistical Analysis: Dependent Variable 2 (Ease of Change Ratings)

```{r}
# loading the data
df2 <- readfile <- read.csv("data/amb_DV2_fin.csv")

# view the data
View(df2)

# Examine the data contents and data types
str(df2)

```

## 1. Cursory EDA 

```{r}
df2 %>%
  group_by(Feature) %>%
  get_summary_stats(DV2, type = "mean_sd")
```

```{r}
bxp2 <- ggboxplot(
  df2, x = "Sonata_Section", y = "DV2",
  color = "Feature", palette = "jco",
  facet.by = "Expertise_Group", short.panel.labs = FALSE
  )
print(bxp2)
```

```{r}
df2 %>%
  group_by(Sonata_Section, Feature, Expertise_Group) %>%
  identify_outliers(DV2)
```

### Assumption of Normality: Shapiro-Wilk Test

Secondly, we can use the Shapiro-Wilk test to examine whether the DV is normally distributed across levels of the IV:

```{r}
df2 %>%
  group_by(Sonata_Section, Feature) %>%
  shapiro_test(DV2)
```

### GG Plot

```{r}
ggqqplot(df2, "DV2", ggtheme = theme_bw()) +
  facet_grid(Sonata_Section ~ Feature , labeller = "label_both")
```
### Assumption of Homogeneity of Variance: Levene Test

```{r}
df2 %>%
  group_by(Sonata_Section, Feature) %>%
  levene_test(DV2 ~ Expertise_Group)
```

## 2. Three-Way Mixed ANOVA: 2*3 Factorial Design (Sonata Section, Feature)

A mixed ANOVA was performed in order to examine the effects of two within subject factors (sonata section and feature) as well as one between subject factor (expertise group) on DV2, ease of change rating.

```{r}
res.aov.bet2 <- anova_test(
  data = df2, dv = DV2, wid = Random_IDs,
  within = c(Sonata_Section, Feature), between=Expertise_Group
)
get_anova_table(res.aov.bet2, correction="GG")
```


```{r}
res.aov.bet2$'Sphericity Corrections'
```

The analysis revealed that there were only main effets of expertise group (p = 0.024) and feature (p = 0.022).

### Examining the Main Effects

Here I will use Tukey HSD tests to examine the main effects. 

**Expertise**

```{r}
df2 %>%
  tukey_hsd(DV2 ~ Expertise_Group, conf.level = 0.95)
```

```{r}
df2 %>%
  group_by(Expertise_Group) %>%
  get_summary_stats(DV2, type = "mean_sd")
```

As was the case with DV1 (ease of hearing ratings), both expert (M = 3.57, sd = 2.06) and intermediate (M = 4.48, sd = 1.75) differed from the novice (M = 1.87, sd = 1.50), indicating that novices struggled to change interpretations. 

**Feature**

```{r}
df2 %>%
  tukey_hsd(DV2 ~ Feature, conf.level = 0.95)

```

```{r}

df2 %>%
  group_by(Feature) %>%
  get_summary_stats(DV2, type = "mean_sd")
```

While the ANOVA revealed a significant effect of feature for DV2, the more conservative nonparametric Tukey HSD test showed that all pair-wise differences of feature were nonsignificant. 


